\newpage
\begin{center}
    \LARGE
    \textbf{\thesistitle}
    
    \vspace{0.5cm}
    
    \authornamefl
    
    \vspace{1cm}
    
    \textbf{Summary}
    
    \vspace{1cm}
\end{center}

Modern fitness and nutrition software applications heavily rely on manual calorie tracking 
and rigid, template-based meal databases. 
These traditional tools lack automated, mathematically precise personalization, requiring
 users to manually balance macro-nutrients through tedious trial and error. 

This thesis introduces FitTrack, an intelligent nutrition management platform designed to
automate custom meal plan generation based on individual physiological requirements.
The system implements a two-stage optimization pipeline: a Multi-Layer Perceptron (MLP)
classifier first shortlists candidate meals by evaluating the user's computed calorie and
protein targets alongside a meal category identifier, and a nonlinear constrained
optimizer then scales each meal's ingredient weights to gram-level precision using
Sequential Least Squares Programming (SLSQP). The optimizer minimizes a weighted
quadratic deviation from the calorie and protein targets, with a minimum fat constraint
and category-specific ingredient bounds ensuring realistic portion compositions.

The application bridges the gap between passive food logging diaries and active, automated
nutritional coaching. Experimental evaluations across six user profiles spanning all
activity groups and dietary goals demonstrate a consistent calorie deviation of under
1.3\% from the computed target, serving as a practical foundation for building
personalized health and wellness applications.

\vspace{0.5cm}
\noindent \textbf{Keywords:} Automated Meal Planning, Nonlinear Optimization, SLSQP, Multi-Layer Perceptron, Personalized Nutrition.