\chapter{Theoretical foundation and bibliographical documentation}
\label{cap:cap2}

\section{Relevant Concepts and Theories}

\subsection{Physiological Caloric and Macro-nutrient Modeling}
Before a software architecture can automate meal generation, it must establish a baseline 
understanding of human energy metabolism. The foundation of personalized nutrition relies 
on two primary metrics: Basal Metabolic Rate (BMR) and Total Daily Energy Expenditure 
(TDEE).

BMR represents the minimal amount of energy, measured in calories, that the human body requires 
to maintain vital vegetative functions—such as cell production, respiration, and temperature 
regulation—at absolute rest in a post-absorptive state \cite[Ch.~2]{techreport:fao:energy}. 
Because BMR accounts for roughly 60\% to 75\% of an individual's daily energy output, it serves 
as the mandatory baseline for any personalized coaching algorithm.

TDEE, on the other hand, represents the total number of calories an individual burns 
over a 24-hour window. It builds upon the BMR by incorporating additional metabolic 
demands, primarily the energy expended during physical exercise, daily functional movement, 
and the thermic effect of food processing, all of which dynamically scale an individual's 
baseline energy requirement based on adult activity vectors \cite[Ch.~5]{techreport:fao:energy}.

To establish a precise nutritional target, the platform must dynamically calculate an 
individual's Total Daily Energy Expenditure (TDEE). This process is executed in a 
two-stage mathematical pipeline: first, determining the Basal Metabolic Rate (BMR), 
and second, applying an activity multiplier coefficient based on the user's lifestyle 
density.

The baseline metabolic requirements are computed using the Mifflin-St Jeor 
equation \cite{Mifflin1990}, which models resting energy expenditure based on 
weight ($w$ in kg), height ($h$ in cm), age ($a$ in years), and a discrete biological sex 
constant ($s$). The mathematical formulation is defined as:

\begin{equation}
BMR = 10w + 6.25h - 5a + s
\end{equation}

where the gender-specific parameter $s$ is defined conditionally as:

\begin{equation}
s = \begin{cases} 
5, & \text{if user is male} \\
-161, & \text{if user is female}
\end{cases}
\end{equation}

To scale the resting metabolic rate to a realistic daily caloric maintenance target, 
the calculated BMR is mapped against standard metabolic activity multipliers derived 
from historical evaluations of the Harris-Benedict principle \cite{Roza1984}. The 
platform defines the final TDEE using an activity factor ($AF$) corresponding to five 
distinct physiological movement groups:

\begin{equation}
TDEE = BMR \times AF
\end{equation}

The activity factor coefficient $AF$ is assigned according to the user's weekly exercise 
intensity vectors:
\begin{align*}
AF &= 1.200 \quad \text{(Inactive: little to no exercise)} \\
AF &= 1.375 \quad \text{(Low: light exercise 1--3 times/week)} \\
AF &= 1.550 \quad \text{(Medium: moderate exercise 3--5 times/week)} \\
AF &= 1.725 \quad \text{(High: intense exercise 6--7 times/week)} \\
AF &= 1.900 \quad \text{(Intense: daily heavy physical labor or athletic sports)}
\end{align*}

\subsection{Multi-Layer Perceptron (MLP) Classifier Foundations}

To automate the selection of recipe baselines within the system, a Multi-Layer Perceptron
(MLP) architecture is utilized as a learned classifier \cite[Ch.~10]{book:geron:handson:2019}.
Instead of executing resource-intensive database queries to scan thousands of potential meal
profiles, the network processes the primary user macro-nutrient requirements instantly to
isolate a highly ranked subset of candidate configurations.

\subsubsection{Structural Topology and Layer Configurations}
The neural network is structured as a feedforward network containing an input layer, two 
fully connected hidden layers, and an unnormalized dense output layer. The input features 
are mapped as a 3-dimensional vector containing the targets for calories, protein, and the 
categorical meal identifier:

\[
x = \begin{bmatrix} \text{Calories} \\ \text{Protein} \\ \text{Meal\_Type\_ID} \end{bmatrix}
\]

The data propagates through the network via successive linear matrix transformations. At each 
layer, the incoming vector is multiplied by a weight matrix ($W$) and combined with a bias 
vector ($b$):

\begin{equation}
\text{Output} = W \cdot x + b
\end{equation}

To introduce non-linear mapping capabilities, the hidden layers apply the Rectified Linear
Unit (ReLU) activation function, which truncates negative values to zero 
\cite[Ch.~10]{book:geron:handson:2019}. The network 
configuration expands the 3-dimensional input into a primary hidden layer of 512 neurons, 
compresses the features into a secondary hidden layer of 256 neurons, and finally projects 
onto the output layer.

\subsubsection{Top-K Output Selection}
The final layer contains a total number of nodes equal to the total number of pre-compiled 
recipe variations in the system registry. Rather than computing an expensive probability 
distribution across all possibilities, the network yields raw structural evaluation scores 
known as logits. 

The application logic extracts the highest-scoring indices from this output vector using a 
Top-K maximum selection operator. This isolates a focused array of candidate recipe IDs 
(where $k=5$ for standard targets or $k=15$ for simple complexity structures) to be passed 
downstream for exact ingredient weight optimization.

\subsection{Nonlinear Constrained Optimization}

To resolve the exact ingredient quantities necessary to fulfill a user's
macro-nutrient targets, the system employs a nonlinear constrained optimizer.
While the Multi-Layer Perceptron classifier selects the most suitable candidate
meal templates, the optimizer determines the precise weight in grams for each
ingredient by solving a continuous constrained minimization problem using the
Sequential Least Squares Programming (SLSQP) method \cite{art:virtanen:scipy:2020}.

A strict Linear Programming formulation was initially considered, which would
express calorie and protein targets as hard equality or inequality constraints.
However, enforcing exact macro-nutrient targets as hard constraints can render
the problem infeasible for certain meal compositions where no combination of
ingredient weights satisfies all constraints simultaneously. To guarantee that
the solver always produces a valid result, the calorie and protein targets are
instead incorporated into a weighted quadratic objective function, allowing the
optimizer to find the closest achievable solution even when exact targets cannot
be met.

\subsubsection{Problem Formulation and the Objective Function}
Each ingredient in the selected meal is treated as a decision variable $x_i$,
representing its mass in grams. The objective minimizes a weighted sum of squared
deviations from the calorie and protein targets:

\begin{equation}
\min \left( \sum_{i=1}^{n} e_i x_i - E_{\text{target}} \right)^2
+ 20 \cdot \left( \sum_{i=1}^{n} p_i x_i - P_{\text{target}} \right)^2
\end{equation}

Where $e_i$ and $p_i$ are the calorie and protein values of ingredient $i$ per
gram respectively, and $E_{\text{target}}$, $P_{\text{target}}$ are the per-meal
calorie and protein targets. The factor of 20 weights the protein deviation more
heavily, reflecting that a gram of protein deviation is more significant to the
nutritional goal than a single calorie deviation.

\subsubsection{Constraints and Boundary Conditions}
A minimum fat inequality constraint was initially formulated to protect metabolic health:

\begin{equation}
\sum_{i=1}^{n} f_i x_i \ge F_{\text{min}}
\end{equation}

Where $f_i$ is the fat content of ingredient $i$ per gram and $F_{\text{min}}$ is
the minimum fat target supplied by the \texttt{DailyPlanService}, calculated as a
per-kilogram value scaled to the user's body weight and dietary goal.

Each decision variable is bounded by category-specific gram limits:

\begin{equation}
x_{\min, i} \le x_i \le x_{\max, i}, \quad \forall i \in \{1, \dots, n\}
\end{equation}

The bounds are assigned per ingredient food category — for example, protein
sources are bounded between 30g and 300g, oils between 5g and 20g, and
vegetables between 30g and 200g. These boundaries encode domain knowledge about
realistic meal portions and prevent the solver from producing nutritionally valid
but culinarily impractical compositions.

\section{Comparative Analysis of Existing Solutions}

The digital nutrition ecosystem features various architectural approaches to 
software-assisted macro-nutrient management. These solutions generally fall into 
two categories: manual logging utilities and rigid algorithmic recipe generators. 

\begin{itemize}
    \item \textbf{Manual Logging Utilities (e.g., MyFitnessPal):} These platforms operate 
    primarily as relational database management interfaces. Users manually search, select, 
    and log food entries post-consumption \cite{url:myfitnesspal}. Caloric guidelines are 
    treated as static targets derived from baseline formulas, placing the operational burden 
    of macro-nutrient mathematical alignment entirely on the end user.
    \item \textbf{Dynamic Metabolic Trackers (e.g., MacroFactor):} These systems use advanced 
    expenditure analytics to continuously recalculate metabolic expenditure based on daily 
    weight and logging inputs \cite{url:macrofactor}. While highly accurate from a 
    physiological modeling perspective, they remain bound to manual data-entry paradigms and 
    lack automated meal generation capabilities.
    \item \textbf{Automated Rule-Based Planners (e.g., Eat This Much):} 
    Standard automated meal planners utilize deterministic decision trees or 
    static search queries to select pre-authored recipes from a fixed 
    index \cite{url:eatthismuch}. These platforms lack micro-granular ingredient 
    flexibility; they cannot dynamically scale individual ingredient quantities 
    to satisfy strict mathematical constraint configurations across an entire day.
\end{itemize}

A technical comparison of these architectural paradigms relative to the proposed system 
is outlined in Table~\ref{tab:comparative_analysis}.

\begin{table}[!htbp]
\centering
\caption{Comparative Matrix of Existing Digital Nutrition Solutions}
\label{tab:comparative_analysis}
\begin{tabular}{|l|c|c|c|c|}
\hline
\textbf{Technical Criterion} & \textbf{MyFitnessPal} & \textbf{MacroFactor} & \textbf{Standard Planners} & \textbf{Proposed System} \\ \hline
Automated Generation        & No                    & No                   & Yes                        & \textbf{Yes}             \\ \hline
ML-Based Meal Pre-Selection & No                    & No                   & No                         & \textbf{Yes (MLP)}       \\ \hline
Exact Gram-Level Scaling    & No                    & No                   & No                         & \textbf{Yes (LP)}        \\ \hline
Dynamic Macro Adjustments   & No                    & Yes                  & No                         & \textbf{Yes}             \\ \hline
Minimal Operational Burden  & No                    & No                   & Moderate                   & \textbf{High}            \\ \hline
\end{tabular}
\end{table}

\section{Identification of Gaps in Current Solutions}

The comparative analysis highlights three primary architectural gaps within the current 
state of the art:

\begin{enumerate}
    \item \textbf{High User Cognitive Load:} Traditional systems depend on continuous 
    manual interaction loops, requiring the user to manually weigh, search, and log 
    individual ingredients post-consumption. This tedious operational cycle quickly 
    induces tracking fatigue, which serves as the primary driver for high user abandonment 
    rates in digital dietetics software.
    \item \textbf{Rigidity of Menu Databases:} Existing automated planners treat recipe 
    structures as static lookups. If an isolated recipe configuration fails to meet a 
    user's calculated macro-nutrient targets exactly, the system cannot modify its internal 
    components. It must either reject the recipe entirely or present an unaligned, 
    mathematically inaccurate menu choice to the user.
    \item \textbf{Lack of Multi-Stage Architecture:} Current software systems treat 
    automated meal planning either as a pure database search or a pure mathematical 
    constraint optimization problem. Applying complex optimization models globally 
    across thousands of raw database rows fails to scale efficiently on standard backend 
    web servers, resulting in substantial processing latencies that degrade the user 
    experience.
\end{enumerate}

\section{Expected System Specifications}

To resolve the identified limitations in existing digital 
dietetics platforms, the proposed system is engineered around a 
derived set of technical and functional specifications:

\begin{itemize}
    \item \textbf{Functional Autonomy:} The platform must automatically synthesize 
    a complete, cohesive daily meal blueprint using calculated target calories, 
    protein thresholds, and minimum fat boundaries. This removes the necessity of 
    manual food combination planning and reduces the user's operational interaction 
    loop to a simple configuration step.
    \item \textbf{Hybrid Pipeline Execution:} The system backend must execute a two-stage 
    processing pipeline to combine performance speed with mathematical precision. 
    A machine learning classification layer (MLP) must handle initial structural baseline
    filtering from the database, which a nonlinear constrained optimizer (SLSQP)
    then scales down to individual ingredient masses in grams.
    \item \textbf{User-Defined Complexity Constraints:} The menu generation pipeline must 
    dynamically adapt candidate food pools based on specified user lifestyle parameters. 
    The system must safely isolate simple, time-efficient recipes containing minimal 
    preparation steps from standard multi-ingredient variants, without causing mathematical 
    failures or violating nutritional safety margins.
\end{itemize}